Dependent Types and Martin-L\"of Type Theory: An Introduction

### 2016-08-03 paper.tex
getting there

\section{Equivalence and Sample Proofs}
In this section, we present the type of Natural numbers, Lists, and the identity
type. To further understand these types, we also include a brief description of
the different forms of equality, namely definitional, judgmental and
propositional. In type theory, the specific differences between the three are
generally misunderstood but are important in understanding basic type theoretic
proof methods, which we also go into some detail in this section with several
examples.

\subsection*{Equality}
%% talk about definitional equality, judgmental equality, prop equality
The first form of equality is much simpler than the others. This form of
equality is called {\it definitional equality} \cite{}. Simply put, definitional
equality is equality through definition, which requires no proof. For example,
we know that for any two natural numbers {\tt m} and {\tt n}, {\tt m `plus` n}
is equal to the sum of {\tt m} and {\tt n} by definition of {\tt `plus`}.

@@

The final form of equivalence is quite a bit more complex than the previous two
and requires a special kind of proof. Propositional equality is equality through
an identity type (e.g. {\tt x ≡ y}). Thus, we now formally introduce the
definition of the identity type:
%% say something that this is the HoTT version of identity
%% possibly also point out that the discrepancy between the two
%% -- "PROPOSITIONAL EQUALITY"
%% -- For a proposition, C, on a proof of x ≡ y for
%% -- any x and y, if we can show that it holds for
%% -- (refl x), then C holds for all proofs, p : x ≡ y.
%% -- "STRICT EQUALITY"
%% -- For a proposition, C, on a reflexive equality (x ≡ x),
%% -- for any x, if we can show that it holds for (refl x), then
%% -- C holds for all other reflexive equalities p : x ≡ x.
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data _≡_ {A : Set} : (a b : A) → Set where
  refl : (a : A) → (a ≡ a)
\end{code}
\end{minipage}
\end{center}
Out of all the types we have introduced, the identity type is the first type
parameterized by terms (viz. {\tt a} and {\tt b}). Until now, we have only
introduced types parameterized by other types. Types parameterized by terms are
classified as an {\em indexed types} \cite{}. Essentially, if we think of types
as shapes (i.e., we can think of them as circles containing points representing
their terms), indexed types change their shape depending on which terms are
passed to them. For example, it is not the case that {\tt y ≡ x} automatically
comes from a proof of {\tt x ≡ y}. This requires a proof of the symmetry of
identity types.

Before we can prove this, however, we must first introduce the recursion and
induction principles of the identity type. Unlike the other types we have
previously introduced, the identity type has two induction principles, both of
which have their own appropriate use case \cite{}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
J : {A : Set} (C : {x y : A} → x ≡ y → Set) →
    (∀ (x : A) → C (refl x)) →
    {x y : A} (p : x ≡ y) → C p
J C c (refl x) = c x

K : {A : Set} (C : {x : A} → x ≡ x → Set) →
    (∀ (x : A) → C (refl x)) →
    {x : A} (p : x ≡ x) → C p
K C c (refl x) = c x

rec≡ : {A : Set} {x y : A} (C : A → Set) →
       (p : x ≡ y) → C x → C y
rec≡ {A} {x} {y} C = 
  J (λ {x} {y} _ → C x → C y)
    (λ x → λ z → z)
    {x} {y} 
\end{code}
\end{minipage}
\end{center}
The two induction principles of the identity type (viz. {\tt J} and {\tt K})
refer to the two ways in which an identity type can be used in a proof, where
{\tt J} is more general. This simply means that {\tt J} can be used for a wider
variety of proofs concerning identity types.

At first glance, the difference between the two induction principles is clear
when looking at the {\tt C} parameter in both. In {\tt J}, {\tt C} is
parameterized with two possibly distinct terms (viz. {\tt x} and {\tt y}),
whereas {\tt K} only mentions one, {\tt x}. This distinction is what allows
{\tt J} to be used for {\em special} identity proofs, that is, in proofs
for identity types of more complex forms of equality (i.e., propositional
equality).

As an example, let's prove the symmetry of identity types using {\tt J}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym_J : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym_J = J ? ?
\end{code}
\end{minipage}
\end{center}
Now, we must fill in both instances of {\tt ?} in the above proof. Similar to
previous proofs, the first of these is our motive and is simply a restatement of
our type statement. In these forms of proofs, we commonly annotate parameterized
identity types with their given equality (viz. {\tt x=y}):
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym_J : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym_J = J (λ {x y} x=y → y ≡ x) ?
\end{code}
\end{minipage}
\end{center}
Now, the last {\tt ?} must be filled in with a proof that our motive holds in
the case that {\tt y} is {\tt x}, that is, we must prove that
{\tt (x : A) → x ≡ x}, which is the type of {\tt refl}. Thus, our completed
proof is the following:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym_J : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym_J = J (λ {x y} x=y → y ≡ x) refl
\end{code}
\end{minipage}
\end{center}

%% \subsection*{Booleans}
%% \begin{center}
%% \begin{minipage}{0.9\textwidth}
%% \begin{code}
%% data 𝔹 : Set where
%%   True  : 𝔹
%%   False : 𝔹

%% rec𝔹 : (C : Set) → C → C → 𝔹 → C
%% rec𝔹 C t f True = t
%% rec𝔹 C t f False = f

%% ind𝔹 : (C : 𝔹 → Set) → C True → C False → (x : 𝔹) → C x
%% ind𝔹 C t f True = t
%% ind𝔹 C t f False = f
%% \end{code}
%% \end{minipage}
%% \end{center}

@@

\section{Discussion}
Say a few things about the paper.

>>>>

\section{Equivalence and Other Types}
In this section, we present the type of Natural numbers, Lists, and the identity
type. To further understand these types, we also include a brief description of
the different forms of equality, namely definitional, judgmental and
propositional. In type theory, the specific differences between the three are
generally misunderstood but are important in understanding basic type theoretic
proof methods, which we also go into some detail in this section with several
examples.

\subsection*{Equality and Identity Types}
%% talk about definitional equality, judgmental equality, prop equality
The first form of equality is much simpler than the others. This form of
equality is called {\it definitional equality} \cite{}. Simply put, definitional
equality is equality through definition, which requires no proof. For example,
we know that for any two natural numbers {\tt m} and {\tt n}, {\tt m `plus` n}
is equal to the sum of {\tt m} and {\tt n} by definition of {\tt `plus`}.

@@

The final form of equivalence is quite a bit more complex than the previous two
and requires a special kind of proof. Propositional equality is equality through
an identity type (e.g. {\tt x ≡ y}). Thus, we now formally introduce the
definition of the identity type.
%% say something that this is the HoTT version of identity
%% possibly also point out that the discrepancy between the two
%% -- "PROPOSITIONAL EQUALITY"
%% -- For a proposition, C, on a proof of x ≡ y for
%% -- any x and y, if we can show that it holds for
%% -- (refl x), then C holds for all proofs, p : x ≡ y.
%% -- "STRICT EQUALITY"
%% -- For a proposition, C, on a reflexive equality (x ≡ x),
%% -- for any x, if we can show that it holds for (refl x), then
%% -- C holds for all other reflexive equalities p : x ≡ x.
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data _≡_ {A : Set} : (a b : A) → Set where
  refl : (a : A) → (a ≡ a)
\end{code}
\end{minipage}
\end{center}
Out of all the types we have introduced, the identity type is the first type
parameterized by terms (viz. {\tt a} and {\tt b}). Until now, we have only
introduced types parameterized by other types. Types parameterized by terms are
classified as an {\em indexed types} \cite{}. Essentially, if we think of types
as shapes (i.e., we can think of them as circles containing points representing
their terms), indexed types change their shape depending on which terms are
passed to them. For example, it is not the case that {\tt y ≡ x} automatically
comes from a proof of {\tt x ≡ y}. This requires a proof of the symmetry of
identity types.

Before we can prove this, however, we must first introduce the recursion and
induction principles of the identity type:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
J : {A : Set} (C : {x y : A} → x ≡ y → Set) →
    (∀ (x : A) → C (refl x)) →
    {x y : A} (p : x ≡ y) → C p
J C c (refl x) = c x

rec≡ : {A : Set} {x y : A} (C : A → Set) →
       (p : x ≡ y) → C x → C y
rec≡ {A} {x} {y} C = 
  J (λ {x} {y} _ → C x → C y)
    (λ x → λ z → z)
    {x} {y} 
\end{code}
\end{minipage}
\end{center}
The more general induction principle for identity types is called {\tt J}.
Later, we discuss a simpler, yet more restricted version. For now, we take
the time to introduce the more general version which is applicable to every
possible identity type \cite{}.

Since identity types are parameterized over two terms, naturally, the motive
parameter (viz. {\tt C}) of their induction principle is also parameterized
by terms (viz. {\tt x} and {\tt y}). The simple reading of {\tt J} is read
similar to other induction principles: given a proposition, {\tt C}, on an
arbitrary identity type, if we can prove that {\tt C} holds for {\tt refl x},
for all {\tt x}, then {\tt C} holds for all identity types. This becomes a bit
more intuitive when we think of what a proof of equality entails: if we want
to prove something for a proof of equality, it suffices to prove it for the
simplest form of equality (i.e., a reflexive equality, {\tt refl}). One can
also note the similarities between {\tt J} and the induction principle for
{\tt ⊤}, since these types both have a single constructor (i.e., {\tt refl}
and {\tt *}, respectively).

Now, let's prove the symmetry of identity types using {\tt J}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym = J ? ?
\end{code}
\end{minipage}
\end{center}
Next, we must fill in both instances of {\tt ?} in the above proof. Similar to
previous proofs, the first of these is our motive and is simply a restatement of
our type statement. In these forms of proofs, we commonly annotate parameterized
identity types with their given equality (viz. {\tt x=y}):
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym = J (λ {x y} x=y → y ≡ x) ?
\end{code}
\end{minipage}
\end{center}
Now, the last {\tt ?} must be filled in with a proof that our motive holds in
the case that {\tt y} is {\tt x}, that is, we must prove that
{\tt (x : A) → x ≡ x}, which is the type of {\tt refl}. Thus, our completed
proof is the following:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym = J (λ {x y} x=y → y ≡ x) refl
\end{code}
\end{minipage}
\end{center}
To reiterate, this is an example of propositional equality. If this were a
definitional or judgmental equality, we would not need to prove symmetry, since
these two forms of equality are automatically reflexive, symmetric, associative,
and transitive. This is the primary and most important difference between other
forms of equality and propositional equality.

A curious fact about {\tt J} is that it does not lend itself to a uniqueness of
identity proof, UIP. In fact, for many reasons, this is actually beneficial and
allows the identity type to be used outside of simple proofs of equality \cite{}.
We discuss this in the final section of this paper and compare {\tt J}
side-by-side with another, less general induction principle for identity types.

\subsection*{Other types}
We now simply provide the remaining types for Natural numbers and lists. With
what we have discussed, these types should not require much explanation.

@@

\section{Exercises}
We have now seen and discussed all the types necessary to start proving more
complex proofs. As an exercise to the reader, we include a list of propositions
to prove. We invite the reader to use only the appropriate induction and
recursion principles in each proof. It helps to first attempt to prove these
propositions using simple pattern matching. As a hint, use a recursion principle
when the proof behaves like a simple function (i.e., defining {\tt `plus`}) and
induction principle when proving propositions {\tt C} for all terms of a given
type (i.e., like {\tt sym}).

\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
-- Identity Types
congruence : {A : Set} {x y : A} (P : A → Set) →
             (p : x ≡ y) → P x → P y
transitivity : {A : Set} {x y z : A} → (x ≡ y) →
               (y ≡ z) → (x ≡ z)

-- Natural numbers (requires the above)
plus : ℕ → ℕ → ℕ
plus-left-unit : (i : ℕ) → plus 0 i ≡ i
plus-right-unit : (i : ℕ) → plus i 0 ≡ i
plus-successor : (i j : ℕ) → plus (suc i) j ≡ plus i (suc j)
plus-commutativity : (i j : ℕ) → plus i j ≡ plus j i
plus-associativity : (x y z : ℕ) →
                     plus x (plus y z) ≡ plus (plus x y) z

-- Lists
_++_ : {A : Set} → List A → List A → List A
reverse : {A : Set} → List A → List A
append-left-unit : {A : Set} (xs : List A) → Nil ++ xs ≡ xs
append-right-unit : {A : Set} (xs : List A) → xs ++ Nil ≡ xs
append-associativity : {A : Set} (xs ys zs : List A) →
                       xs ++ (ys ++ zs) ≡ (xs ++ ys) ++ zs
append-distributivity : {A : Set} (xs ys : List A) →
                        reverse (xs ++ ys) ≡ (reverse ys) ++ (reverse xs) 
double-reverse : {A : Set} (xs : List A) → reverse (reverse xs) ≡ xs

\end{code}
\end{minipage}
\end{center}
